---
title: "Developing the Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developing the Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Once your repository is initialized (see [Launching a Picard Study](launching_a_study.html)), the next phase is to develop your analysis pipeline. This vignette walks you through the complete development workflow on the `develop` branch, from defining inputs through testing your code.

## Development Workflow Overview

The typical workflow is:

1. Finalize Git and renv setup
2. Create your development branch
3. Define your inputs (cohorts and concept sets)
4. Create analysis tasks and supporting code
5. Commit your changes with `saveWork()`
6. Test individual tasks and the full pipeline
7. Iterate until your analysis is complete

## Step 1: Finalize Git and renv Setup

**Git:** You already have a local repository. If you provided `gitRemote` during setup, your code is already synced. If not, see [Setting Up Git Version Control](launching_a_study.html#setting-up-git-version-control) in the launching vignette to push your code to a remote.

**renv:** This is your responsibility. It's **highly encouraged** to set this up immediately:

```{r eval = FALSE}
renv::init(project = "~/studies/diabetes_study")
```

This captures your current R environment, ensuring reproducibility for your team.

## Step 2: Create Your Development Branch

All development work happens on the `develop` branch (never on `main`):

```bash
# Create and switch to develop branch
git checkout -b develop
git push -u origin develop
```

All subsequent work will be done here and tested before any production execution.

## Step 3: Define Your Inputs (Cohorts and Concept Sets)

Before writing analysis code, you need to define the populations and phenotypes. This happens in the `inputs/` folder:

**For Cohorts:** Use the interactive editor to add your study populations:

```{r eval = FALSE}
launchCohortsLoadEditor(cohortsFolderPath = here::here("inputs/cohorts"))
```

This opens a Shiny app where you can add cohort metadata without editing CSV files directly.

**For Concept Sets:** Similarly, define your phenotypes:

```{r eval = FALSE}
launchConceptSetsLoadEditor(conceptSetsFolderPath = here::here("inputs/conceptSets"))
```

See [Loading Inputs](loading_inputs.html) for detailed guidance on importing from ATLAS and managing manifests.

## Step 4: Create Analysis Tasks and Supporting Code

Analysis happens in the `analysis/` folder, which has two key subdirectories:

**Tasks** (`analysis/tasks/`) - Main workflow files executed by the pipeline
**Source** (`analysis/src/`) - Supporting functions and utilities used by tasks

### Creating a Task

Use `makeTaskFile()` to scaffold a new task:

```{r eval = FALSE}
makeTaskFile(
  nameOfTask = "01_DescriptiveStats",
  author = "Jane Doe",
  description = "Generate descriptive statistics for primary cohort",
  projectPath = here::here(),
  openFile = TRUE
)
```

This creates a standardized R script in `analysis/tasks/01_DescriptiveStats.R` with a template structure.

**Task Naming Convention:** Use `NN_taskName.R` format where NN is a two-digit number (01, 02, 03). Tasks execute in alphabetical order, so the numbering controls execution sequence.

### Creating Supporting Source Files

For reusable utility functions, create files in `analysis/src/` using `makeSrcFile()`:

```{r eval = FALSE}
makeSrcFile(
  fileName = "custom_analysis_functions",
  author = "Jane Doe",
  description = "Helper functions for cohort calculations and data validation"
)
```

This creates a standardized R script in `analysis/src/custom_analysis_functions.R` with a template structure for your utility functions.

**Important:** Treat source files like R package development:

- **Use `package::function()` notation** instead of `library()` calls:
  ```{r eval = FALSE}
  # Good: Use explicit namespace
  result <- dplyr::filter(data, age > 18)
  
  # Avoid: Loading entire library
  # library(dplyr)
  # result <- filter(data, age > 18)
  ```
  
  This makes dependencies explicit and avoids namespace conflicts.

- **Document each function** with a clear comment block describing:
  - What the function does
  - What inputs it expects
  - What it returns
  
  Example:
  ```{r eval = FALSE}
  # Purpose: Calculate age in years from birth date
  # Inputs: birth_date (Date or character in YYYY-MM-DD format)
  # Returns: Numeric age in years
  calculate_age <- function(birth_date) {
    birth_date <- as.Date(birth_date)
    age <- as.numeric(lubridate::interval(birth_date, Sys.Date()), "years")
    return(floor(age))
  }
  ```

Then source these files from your tasks:

```{r eval = FALSE}
source(here::here("analysis/src/custom_analysis_functions.R"))
```

### Creating SqlRender SQL Queries

For parameterized SQL queries used with SqlRender, create files in `analysis/src/sql/` using `makeSrcSqlFile()`:

```{r eval = FALSE}
makeSrcSqlFile(
  fileName = "condition_occurrence_summary",
  author = "Jane Doe",
  description = "Count conditions by month in the CDM"
)
```

This creates a standardized SQL file in `analysis/src/sql/condition_occurrence_summary.sql` with placeholders for SqlRender parameters.

**Important:** SqlRender uses `@paramName` notation for substitution:

```sql
/* Document your parameters at the top of the file
   @cdmDatabaseSchema - Schema containing OMOP CDM tables
   @workDatabaseSchema - Schema for output tables
*/

SELECT *
FROM @cdmDatabaseSchema.condition_occurrence
WHERE month_of_year = 6
LIMIT 100;
```

Then render and execute from your R tasks:

```{r eval = FALSE}
# Read the SQL file
sql <- readr::read_file(here::here("analysis/src/sql/condition_occurrence_summary.sql"))

# Render parameters using SqlRender
rendered_sql <- SqlRender::render(
  sql,
  cdmDatabaseSchema = "omop_cdm",
  workDatabaseSchema = "work_schema"
)

# Execute the rendered query
result <- DatabaseConnector::executeSql(connection, rendered_sql)
```

## Step 5: Commit Changes with `saveWork()`

After creating your analysis code, commit your changes to track them in Git. Picard provides `saveWork()` to make Git commits easy for new users:

```{r eval = FALSE}
saveWork("Add descriptive statistics task")
```

`saveWork()` is a convenience wrapper around Git operations that handles the common workflow for you:

1. **Show what changed** - Displays all modified, new, and deleted files
2. **Confirm commit** - Asks you to approve the changes before proceeding
3. **Stage files** - Automatically stages all changes
4. **Commit** - Creates a commit with your message
5. **Pull from remote** - Gets the latest changes from your team
6. **Push to remote** - Syncs your commit to the shared repository

**Key features:**

- **Protection from mistakes:** Prevents you from accidentally committing to the `main` branch
- **Interactive confirmation:** Shows exactly which files changed and asks you to confirm before committing
- **Work-in-progress commits:** Use `push = FALSE` to commit locally without pushing (useful for saving work mid-session):
  
```{r eval = FALSE}
saveWork("WIP: Testing new validation logic", push = FALSE)
```

- **Automatic branch creation:** If your feature branch doesn't exist, `saveWork()` creates it for you
- **Syncs with team:** Pulls remote changes before pushing your work, reducing merge conflicts

**Full function signature:**

```{r eval = FALSE}
saveWork(
  commitMessage,                    # Required: Descriptive message
  branch = get_current_branch(),    # Current branch (or specify new one)
  push = TRUE,                      # FALSE for local-only commits
  gitRemoteName = "origin"          # Remote repository name
)
```

**For new Git users:** Use `saveWork()` regularly as you develop. It simplifies the entire commit workflow and protects you from common Git mistakes.

**For advanced users:** You can use Git directly if you prefer. Many developers commit changes as they go using:

```bash
git add analysis/tasks/01_DescriptiveStats.R
git commit -m "Add descriptive statistics task"
git pull origin develop
git push origin develop
```

Either approach works—use what's comfortable for your workflow.

## Step 6: Test Your Development Work

Picard provides two testing modalities on the `develop` branch:

### Test Single Task with `testStudyTask()`

Test the current task file in isolation:

```{r eval = FALSE}
# Open your task file (e.g., 01_DescriptiveStats.R) and run:
testStudyTask()
```

This executes only the current file, allowing rapid iteration on specific tasks without running the entire pipeline.

### Test Full Pipeline with `testStudyPipeline()`

Once individual tasks work, test the complete pipeline:

```{r eval = FALSE}
testStudyPipeline()
```

This runs all tasks in sequence (in alphabetical order) with the same configuration as production, helping you catch integration issues.

**Testing Workflow:**
1. Edit a task file
2. Run `testStudyTask()` to verify the individual task works
3. Once satisfied, run `testStudyPipeline()` to verify the full workflow
4. Commit your changes to `develop`

### Commit and Push to Develop

After testing successfully:

```bash
git add analysis/tasks/01_DescriptiveStats.R
git add analysis/src/custom_analysis_functions.R
git commit -m "Add descriptive stats task and utility functions"
git push origin develop
```

## Next Steps: Production Execution

Once your analysis is tested and validated on `develop`, see [Running the Pipeline](running_the_pipeline.html) for instructions on production execution, which involves creating a release branch and running the full validated pipeline.

## Quick Reference: Key Functions by Phase

| Phase | Function | Purpose |
|-------|----------|---------|
| **Setup** | `launchCohortsLoadEditor()` | Interactive cohort metadata editor |
| **Setup** | `launchConceptSetsLoadEditor()` | Interactive concept set editor |
| **Development** | `makeTaskFile()` | Create new analysis task |
| **Development** | `makeSrcFile()` | (optional) Create utility/helper function file |
| **Development** | `makeSrcSqlFile()` | (optional) Create parameterized SQL query file |
| **Development** | `saveWork()` | Commit and push code changes to remote |
| **Testing** | `testStudyTask()` | Test current task file only |
| **Testing** | `testStudyPipeline()` | Test full pipeline end-to-end |
| **Production** | See Running the Pipeline | Execute on release branch |

## See Also

- [Launching a Picard Study](launching_a_study.html) - Repository initialization and setup
- [Loading Inputs](loading_inputs.html) - Setting up cohorts and concept sets
- [Running the Pipeline](running_the_pipeline.html) - Production pipeline execution
