---
title: "Post-Processing Steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-Processing Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

After your pipeline runs successfully via `testStudyPipeline()` or `execStudyPipeline()`, raw analytical outputs are saved in `exec/results/[database]/[version]/`. Post-processing transforms these raw files into polished results ready for dissemination, visualization, and publication.

Picard post-processing includes:

- **Data formatting:** Convert CSV outputs to analysis-ready formats (xlsx, parquet)
- **Result aggregation:** Combine results across databases for comparison
- **Visualization:** Generate plots and tables for reports
- **Quality assurance:** Check for anomalies and data issues
- **Documentation:** Create vignettes and reports explaining methods and results
- **Archival:** Store important results with metadata for long-term access

This vignette walks through the post-processing workflow.

## Understanding Post-Processing in Picard

### Raw vs. Processed Results

**Raw Results** (in `exec/results/[db]/[version]/`):
- CSVs and data files output directly by analysis tasks
- Intermediate calculations (diagnostics, intermediate statistics)
- Unformatted, potentially large, analysis-ready format

**Processed Results** (in `dissemination/`):
- Finalized tables ready for publication
- Formatted xlsx files with styling
- Visualizations and graphics
- Summary reports combining multiple analyses

### Workflow Phases

**Phase 1: Immediate Post-Processing**
After production pipeline completes, move and consolidate results:

```
exec/results/[db]/[version]/ → dissemination/results/[analysis]/
```

**Phase 2: Quality Assurance**
Review raw results for errors, anomalies, missing data

**Phase 3: Formatting & Aggregation**
Convert to publication-ready formats, combine across databases

**Phase 4: Visualization**
Create figures, tables, and interactive elements

**Phase 5: Documentation**
Write vignettes and technical reports

## Phase 1: Moving and Consolidating Results

### Organize Results by Analysis Type

Create subdirectories in `dissemination/results/` that mirror your analytical structure:

```
dissemination/results/
├── 01_descriptive/
│   ├── demographics.csv
│   ├── flowchart.csv
│   └── sequenceOfEvents.csv
├── 02_primary/
│   ├── mainComparisonResults.csv
│   ├── estimatedEffects.csv
│   └── confidenceIntervals.csv
├── 03_sensitivity/
│   ├── ageSubgroup.csv
│   ├── sexSubgroup.csv
│   └── robustnessTest.csv
└── 04_diagnostics/
    ├── cohortDiagnostics.csv
    └── propensityScoreAnalysis.csv
```

### Consolidate Multi-Database Results

If you ran on multiple databases, create a consolidated summary:

```{r eval = FALSE}
library(picard)
library(dplyr)
library(readr)

# Read results from each database
primary_results <- read_csv("exec/results/primary_db/1.0.0/02_primary/mainResults.csv")
secondary_results <- read_csv("exec/results/secondary_db/1.0.0/02_primary/mainResults.csv")

# Add database identifier
primary_results <- primary_results %>% mutate(database = "Primary DB")
secondary_results <- secondary_results %>% mutate(database = "Secondary DB")

# Combine
combined_results <- bind_rows(primary_results, secondary_results)

# Save consolidated
write_csv(combined_results, "dissemination/results/02_primary/combined_mainResults.csv")
```

## Phase 2: Quality Assurance

### Check for Missing Values

```{r eval = FALSE}
# Load results
results <- read_csv("dissemination/results/02_primary/mainResults.csv")

# Check for missing values
missing_summary <- results %>%
  summarise(across(everything(), ~sum(is.na(.))))

print(missing_summary)

# Flag if unexpected missingness
if (any(missing_summary > 0)) {
  warning("Found missing values in results. Investigate before publication.")
}
```

### Validate Numeric Ranges

```{r eval = FALSE}
# Check for impossible values (e.g., proportions > 1, negative cell counts)
results <- results %>%
  mutate(
    proportion_valid = proportion >= 0 & proportion <= 1,
    count_valid = count >= 0,
    ci_valid = ci_lower <= estimate & estimate <= ci_upper
  )

# Report any issues
if (!all(results$proportion_valid)) {
  warning("Found invalid proportions outside [0,1]")
}
```

### Review Numeric Precision

```{r eval = FALSE}
# Check if numbers have appropriate precision
results %>%
  select(where(is.numeric)) %>%
  summarize(across(everything(), list(
    min = ~min(.),
    max = ~max(.),
    mean = ~mean(.),
    na_count = ~sum(is.na(.))
  )))
```

### Consistency Checks

```{r eval = FALSE}
# Ensure row counts match across related results
descriptive_count <- nrow(read_csv("dissemination/results/01_descriptive/demographics.csv"))
primary_count <- nrow(read_csv("dissemination/results/02_primary/mainResults.csv"))

if (descriptive_count != primary_count) {
  warning(glue::glue(
    "Row count mismatch: descriptive ({descriptive_count}) vs primary ({primary_count})"
  ))
}
```

## Phase 3: Formatting and Aggregation

### Convert to Excel with Formatting

```{r eval = FALSE}
library(openxlsx)

# Read results
results <- read_csv("dissemination/results/02_primary/mainResults.csv")

# Create Excel workbook
wb <- createWorkbook()
addWorksheet(wb, "Results")

# Write data
writeData(wb, "Results", results)

# Format header row
headerStyle <- createStyle(
  fgFill = "#4472C4",
  fontColour = "white",
  textDecoration = "bold"
)
addStyle(wb, "Results", headerStyle, rows = 1, cols = 1:ncol(results))

# Adjust column widths
setColWidths(wb, "Results", cols = 1:ncol(results), widths = "auto")

# Save
saveWorkbook(wb, "dissemination/export/pretty/mainResults.xlsx", overwrite = TRUE)
```

### Summarize Results Across Databases

```{r eval = FALSE}
library(dplyr)

# Read results from all databases
databases <- c("primary_db", "secondary_db", "tertiary_db")

all_results <- map_df(databases, ~{
  db_name <- .x
  read_csv(glue::glue("exec/results/{db_name}/1.0.0/02_primary/mainResults.csv")) %>%
    mutate(database = db_name)
})

# Calculate summary statistics
summary <- all_results %>%
  group_by(analysis) %>%
  summarize(
    n_databases = n_distinct(database),
    estimate_mean = mean(estimate),
    estimate_sd = sd(estimate),
    .groups = "drop"
  )

write_csv(summary, "dissemination/results/02_primary/summary_acrossDatabases.csv")
```

### Create Result Comparison Tables

```{r eval = FALSE}
library(dplyr)
library(tidyr)

# Read multi-database results
combined <- read_csv("dissemination/results/02_primary/combined_mainResults.csv")

# Create wide format for comparison
comparison <- combined %>%
  select(analysis, estimate, ci_lower, ci_upper, database) %>%
  pivot_wider(
    names_from = database,
    values_from = c(estimate, ci_lower, ci_upper)
  )

write_csv(comparison, "dissemination/results/02_primary/comparison_acrossDB.csv")
```

## Phase 4: Visualization

### Generate Descriptive Plots

```{r eval = FALSE}
library(ggplot2)

# Load data
demographics <- read_csv("dissemination/results/01_descriptive/demographics.csv")

# Plot age distribution
p <- ggplot(demographics, aes(x = age_group, y = percentage)) +
  geom_col(fill = "#4472C4") +
  labs(
    title = "Age Distribution of Study Population",
    x = "Age Group",
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("dissemination/results/01_descriptive/ageDistribution.png", p, width = 8, height = 5)
```

### Create Forest Plots for Effect Estimates

```{r eval = FALSE}
library(ggplot2)

# Load results
results <- read_csv("dissemination/results/02_primary/mainResults.csv")

# Create forest plot
p <- ggplot(results, aes(y = contrast, x = estimate)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(
    title = "Primary Analysis Results",
    x = "Hazard Ratio (95% CI)",
    y = "Comparison"
  ) +
  scale_x_log10() +
  theme_minimal()

ggsave("dissemination/results/02_primary/forestPlot.png", p, width = 8, height = 6)
```

### Flowchart Documentation

```{r eval = FALSE}
# Flowchart data from cohort generation
flowchart <- tribble(
  ~step, ~n,
  "Initial cohort", 1234567,
  "Excluded: Age <18", 234567,
  "Excluded: No prior dx", 123456,
  "Primary population", 876000,
  "Exposed group", 456000,
  "Unexposed group", 420000
)

write_csv(flowchart, "dissemination/results/01_descriptive/flowchart.csv")
```

## Phase 5: Documentation

### Create Quarto/R Markdown Reports

```{r eval = FALSE}
# Create dissemination/quarto/mainReport.qmd

# Report template structure:
# - Executive summary
# - Methods (cohort definitions, analyses)
# - Results (tables, figures)
# - Interpretation
# - Appendices (detailed methods, sensitivity analyses)
```

Create a Quarto file that renders results into a professional report:

```bash
# In dissemination/quarto/mainReport.qmd

---
title: "Study Results Report"
author: "Research Team"
date: "`r Sys.Date()`"
format: html
---

## Executive Summary

[Brief summary of key findings]

## Methods

### Study Population

[Describe cohort definitions from manifests]

### Analysis

[Describe analytical approaches]

## Results

```{r}
# Load and display results
results <- read_csv("../results/02_primary/mainResults.csv")
knitr::kable(results)
```

## Appendices

[Sensitivity analyses, diagnostic results, etc.]
```

Render with:

```{r eval = FALSE}
quarto::quarto_render("dissemination/quarto/mainReport.qmd")
```

### Document Methods and Assumptions

Create a methods document explaining your analyses:

```
dissemination/documents/methods.md

## Cohort Definitions

- **Primary Population:** Type 2 Diabetes (see inputs/cohorts/cohortManifest.sqlite ID 1)
- **Exposure:** Antidiabetic Medication (see inputs/conceptSets/conceptSetManifest.sqlite ID 1)
- **Outcome:** Acute MI (see inputs/conceptSets/conceptSetManifest.sqlite ID 2)

## Study Design

[Comparative cohort design, XXX-XXXXX patient-years, ...]

## Statistical Methods

[Analysis approach, adjustment variables, sensitivity tests]
```

## Phase 6: Archival

### version 1Archive Results

```{r eval = FALSE}
library(picard)

# Archive main results
zipAndArchive(input = "exportPretty")

# Archive analysis
zipAndArchive(input = "exportMerge")
```

This creates timestamped archives in `exec/archive/`:

```
exec/archive/
├── export_pretty_1.0.0_20260407_093045.zip
├── export_merge_1.0.0_20260407_093045.zip
└── quarto_site_1.0.0_20260407_093045.zip
```

### Metadata for Archives

Create a README in `exec/archive/` documenting your archives:

```
# Archive Manifest

## Version 1.0.0 (April 7, 2026)

- Filename: export_pretty_1.0.0_20260407_093045.zip
- Size: 245 MB
- Content: Formatted xlsx results and analysis output
- Databases: Primary DB, Secondary DB
- Analysis tasks: 01-05 complete
- Notes: Initial publication version
```

## Integration with Study Pipeline

Post-processing is typically performed immediately after production pipeline completion:

```
1. Commit latest changes to develop branch
2. Run: source("main.R")
   [Pipeline runs and completes]
3. Review PENDING_PR.md
4. Run post-processing tasks
5. Commit formatted results to dissemination/
6. Create PR with both pipeline + processed results
```

## Best Practices

### Reproducibility

All post-processing scripts should be version-controlled:

```
analysis/src/
├── postProcessing.R      # Main post-processing functions
├── formatting.R          # Excel formatting helpers
└── visualization.R       # Common graphing functions
```

Reference these from your post-processing workflows.

### Documentation

Document assumptions and cleaning steps:

```r
# Remove rows with missing exposure data
results_clean <- results %>%
  filter(!is.na(exposure_status))  # Removed N rows

# Flag outliers for review
results_clean <- results_clean %>%
  mutate(
    is_outlier = abs(scale(estimate)) > 3,
    note = if_else(is_outlier, "Review before publication", NA_character_)
  )
```

### Quality Gates

Establish criteria before publishing:

```r
# Quality assurance checklist
qa_checks <- list(
  no_missing = all(!is.na(results$estimate)),
  valid_ci = all(results$ci_lower <= results$estimate & 
                 results$estimate <= results$ci_upper),
  databases_match = n_distinct(results$database) == expected_databases,
  row_counts_match = nrow(results) == expected_rows
)

# Only proceed if all pass
if (!all(unlist(qa_checks))) {
  stop("Quality assurance checks failed. Review results before publication.")
}
```

## Troubleshooting

### "Results folder not found"

Check that pipeline completed successfully:

```{r eval = FALSE}
# Verify results exist
fs::dir_exists("exec/results/primary_db/1.0.0/02_primary/")
```

### "Merging results gives mismatched rows"

Ensure databases used consistent cohort definitions:

```{r eval = FALSE}
# Check cohort counts matched
primary_counts <- read_csv("exec/results/primary_db/1.0.0/00_buildCohorts/cohortCounts.csv")
secondary_counts <- read_csv("exec/results/secondary_db/1.0.0/00_buildCohorts/cohortCounts.csv")

# Compare
setdiff(primary_counts$id, secondary_counts$id)  # If empty, all IDs match
```

### "xlsx writing failed"

Ensure openxlsx is installed:

```{r eval = FALSE}
install.packages("openxlsx")
```

## Next Steps

1. **Immediate:** Run post-processing after each production pipeline execution
2. **Ongoing:** Create reusable post-processing scripts in analysis/src/
3. **Final:** Archive results and create final report for dissemination

## See Also

- [Running the Pipeline](running_the_pipeline.html) - Where results come from
- [The Ulysses Standard Repository Structure](picard_repository_structure.html) - Where results go
- [Evidence Generation Plan](evidence_generation_plan.html) - Strategic organization of outputs
