---
title: "Running the Pipeline: Test vs. Production Execution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the Pipeline: Test vs. Production Execution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Picard distinguishes between two execution modes: **test mode** for rapid iteration during development, and **production mode** for formal analysis runs. This separation ensures code quality while enabling rapid experimentation.

- **Test Mode:** Fast, forgiving iteration during development
- **Production Mode:** Rigorous validation, versioning, and audit trails for official results

This vignette explains when to use each mode, how they differ, and how to run both.

## The Two Pipelines

When you launch a Picard study, two execution scripts are created:

### main.R — Production Pipeline

Located in the project root, `main.R` is the official execution script. It:

- Validates your code state (git clean, all changes committed)
- Increments your study version (semantic versioning)
- Runs the complete pipeline with all validations
- Creates a release branch for reproducibility
- Generates PR metadata for code review
- Saves production-quality results

```{r eval = FALSE}
# Run production pipeline
source("main.R")
```

### extras/test_main.R — Development Pipeline

Located in `extras/`, `test_main.R` is the development variant. It:

- Skips git and code validation
- Uses a "dev" version folder (not production-versioned)
- Allows rapid iteration without version management overhead
- Provides immediate visual feedback
- Ideal for exploratory analysis and task debugging

```{r eval = FALSE}
# Run development pipeline
source("extras/test_main.R")
```

## When to Use Test Mode

Use test mode (`extras/test_main.R`) for:

- **Developing new analysis tasks:** Write, test, iterate on individual tasks
- **Exploring data:** Quick diagnostic runs to understand your database
- **Debugging issues:** Rapid feedback on analysis errors
- **Team collaboration on features:** Before committing to production
- **Any work that's not a formal analysis run**

Test mode places results in `exec/results/[database]/dev/` so they don't interfere with versioned results.

Example workflow:

```{r eval = FALSE}
# 1. Create a new task file in analysis/tasks/04_newAnalysis.R
# 2. Open extras/test_main.R and modify config as needed
source("extras/test_main.R")

# 3. Check output in exec/results/[database]/dev/
# 4. Results are there—iterate and refine
# 5. Once satisfied, commit code and prepare for production run
```

## When to Use Production Mode

Use production mode (`main.R`) for:

- **Formal analysis runs:** Official results for publications or regulatory submissions
- **Final results:** When you're confident in the code and ready for version history
- **Studyathon submissions:** Required for most OHDSI initiatives
- **Multi-database comparisons:** Ensures consistency across databases
- **Code review:** Results from `main.R` go through PR review before being accepted

Production mode places versioned results in `exec/results/[database]/[version]/` (e.g., `1.0.0/`).

## Differences Between Test and Production

| Aspect | Test Mode | Production Mode |
|--------|-----------|-----------------|
| **Script** | extras/test_main.R | main.R |
| **Version** | Always "dev" | Semantic version (1.0.0, etc.) |
| **Git validation** | Skipped | Required (branch, clean state) |
| **Code validation** | Skipped | Full validation |
| **renv snapshot** | Skipped | Captured for reproducibility |
| **Output folder** | `[db]/dev/` | `[db]/[version]/` |
| **PR workflow** | N/A | Automatic PR metadata |
| **Iteration speed** | Very fast | Slower (validation overhead) |
| **Use case** | Development | Official results |

## Running Test Mode

### Basic Usage

```{r eval = FALSE}
# Navigate to your study repository
setwd("~/studies/myStudy")

# Run test pipeline on develop branch
source("extras/test_main.R")
```

This executes all tasks in `analysis/tasks/` with configuration from the first config block.

### Test Individual Tasks

Instead of running the full pipeline, test a single task during development:

```{r eval = FALSE}
library(picard)

# Test just the descriptive statistics task
testStudyTask(
  taskFile = "02_descriptiveStats.R",
  configBlock = "primaryDB"
)
```

Results appear in `exec/results/primary_db/dev/02_descriptiveStats/`.

### Customize Test Configuration

The `extras/test_main.R` script is a template you can modify:

```{r eval = FALSE}
# In extras/test_main.R, you can customize:

# Run only specific config blocks (instead of all)
configBlocks <- "primaryDB"  # or c("primaryDB", "secondaryDB")

# Run only specific tasks by editing the script
# Or use testStudyTask() for individual tasks

# Set custom output messages
verbose <- TRUE

# Then source the script
source("extras/test_main.R")
```

## Running Production Mode

### Prerequisites

Before running production mode:

1. **Commit all changes:** `git add .` and `git commit -m "..."`
2. **Be on develop branch (or feature branch):** `git checkout develop`
3. **Pull latest changes:** `git pull`
4. **Verify configuration:** Check config.yml for correctness

### Basic Usage

```{r eval = FALSE}
# Navigate to study repository
setwd("~/studies/myStudy")

# Run production pipeline with patch version increment
source("main.R")

# When prompted, answer questions about version increment:
# What type of version change? [major/minor/patch]
# You typically choose: "patch" (bug fixes), "minor" (new analyses), "major" (breaking changes)
```

### Programmatic Production Execution

```{r eval = FALSE}
library(picard)

# Run production pipeline directly
execStudyPipeline(
  configBlock = c("primaryDB", "secondaryDB"),
  updateType = "minor"  # Version increment type
)
```

### Version Increment Types

Choose the appropriate semantic version increment:

- **PATCH (1.0.0 → 1.0.1):** Bug fixes, data corrections, no new analyses
- **MINOR (1.0.0 → 1.1.0):** New analyses or features added (backward compatible)
- **MAJOR (1.0.0 → 2.0.0):** Breaking changes or study redesign

## Understanding the Pipeline Workflow

Both pipelines follow the same general workflow:

### Phase 1: Setup

```r
Validate configuration
Load execution settings
Create output directories
```

### Phase 2: Generate Cohorts

```r
Load cohort and concept set manifests
Validate all definitions exist
Generate cohorts in database
Retrieve cohort counts
```

### Phase 3: Run Analysis Tasks

For each task in `analysis/tasks/`:
```r
Load configuration
Execute task code
Check for errors
Record results
```

### Phase 4: Post-Processing (Production Only)

```r
Generate version logs
Create PR metadata
Save PENDING_PR.md
```

## Handling Errors and Failures

### Test Mode Errors

If a task fails in test mode, the pipeline stops and displays the error:

```
Error in task 02_descriptiveStats.R:
object 'cohortData' not found
```

Fix the error in your task file and re-run:

```{r eval = FALSE}
source("extras/test_main.R")
```

### Production Mode Errors

Production mode validates more strictly. Common issues:

**"Cannot run production pipeline on main branch!"**
- Solution: Switch to develop: `git checkout develop`

**"Code state validation failed - uncommitted changes"**
- Solution: Commit all changes: `git add .` and `git commit -m "..."`

**"Cohort manifest not found"**
- Solution: See [Loading Inputs](loading_inputs.html)

## Reviewing Results

### Test Mode Results

After test mode, review results in:

```
exec/results/[database]/dev/
├── 00_buildCohorts/          # Cohort counts
├── 01_firstAnalysisTask/     # Task 1 outputs
├── 02_secondAnalysisTask/    # Task 2 outputs
└── picard_log_dev_*.txt      # Execution log
```

### Production Mode Results

After production mode, results are in versioned folders:

```
exec/results/[database]/1.1.0/       # Version 1.1.0
├── 00_buildCohorts/
├── 01_firstAnalysisTask/
├── 02_secondAnalysisTask/
└── picard_log_1.1.0_*.txt
```

Plus additional files for code review:

```
PENDING_PR.md                       # PR details for manual review
NEWS.md                              # Updated with version info
```

## Code Review Workflow

Production mode enables structured code review:

1. **Run pipeline:** `source("main.R")` on develop branch
2. **Review PENDING_PR.md:** Check proposed version, changes logged in NEWS.md
3. **Review code:** Inspect changes on release branch: `git checkout release/1.1.0`
4. **Create PR:** Use details from PENDING_PR.md to create PR in GitHub/Bitbucket
5. **Merge:** After review and approval, merge to main
6. **Cleanup:** Run `clearPendingPR()` to remove metadata file

## Integration with Git

### Git Branches

Test mode works on any branch. Production mode:

1. Creates a release branch: `release/[version]`
2. Runs pipeline on that branch
3. Saves PR metadata pointing to main
4. Expects manual PR creation and merge

```
main ←──── PR from release/1.1.0 ─── release/1.1.0
  ↑                                        ↑
  │                                        └─ Production run here
  │                                          (all commits included)
  └────────────────────────────────────────── Merged after review
```

### Version Tags

After merging to main, create a git tag for the version:

```{r eval = FALSE}
# After PR is merged to main
git tag -a v1.1.0 -m "Release version 1.1.0"
git push origin v1.1.0
```

## Monitoring Pipeline Execution

### Log Files

Picard creates detailed execution logs:

- **test_main.R run:** `exec/logs/picard_log_dev_*.txt`
- **main.R run:** `exec/logs/picard_log_1.1.0_*.txt`

Review logs to understand which tasks ran, their duration, and any warnings:

```
[14:32:01] Starting cohort generation...
[14:32:15] Cohort generation completed successfully!
[14:32:16] Executing task 1/3: 01_descriptiveStats.R
[14:33:42] ✓ Task completed successfully
[14:33:43] Executing task 2/3: 02_primaryAnalysis.R
```

### Cohort Counts

After any pipeline run, check `00_buildCohorts/cohortCounts.csv` to verify cohorts were generated:

```
id,label,cohort_entries,cohort_subjects
1,Type 2 Diabetes,245897,123456
2,CVD Comparator,189234,98765
3,MI Outcome,34567,12345
```

## Troubleshooting

### "Tasks not running in expected order"

Picard runs tasks in alphabetical order. Ensure file names have numeric prefixes:

```
01_buildCohorts.R          ✓ Runs first
02_descriptiveAnalysis.R   ✓ Runs second
03_primaryAnalysis.R       ✓ Runs third
analysis_task.R            ✗ Runs last (no prefix)
```

### "Results folder not created"

Manually create output folder:

```{r eval = FALSE}
# Ensure output structure exists
exec_path <- fs::path(here::here(), "exec/results/primary_db/dev")
fs::dir_create(exec_path, recurse = TRUE)
```

### "Previous version results disappeared"

Results are organized by version in `exec/results/[database]/[version]/`. Check different version folders:

```{r eval = FALSE}
# List all version folders
list.dirs("exec/results/primary_db", recursive = FALSE)
```

## Next Steps

1. **Start with test mode:** Develop and refine your tasks using `extras/test_main.R`
2. **Verify code quality:** Ensure tasks run successfully multiple times
3. **Switch to production:** When ready for official results, use `main.R`
4. **Archive results:** Use `zipAndArchive()` to preserve important results

## See Also

- [The Ulysses Standard Repository Structure](picard_repository_structure.html) - Where results are organized
- [Launching a Study](launching_a_study.html) - Initial setup
- [Loading Inputs](loading_inputs.html) - Cohort and concept set setup
- [Post-Processing Steps](post_processing.html) - Working with results after execution
